Trauma treatment delay

### 2023-12-19 manuscriptlongversion.Rmd
seminar version

Trauma, defined as the clinical entity composed of physical injury and
the body\'s associated response (1), is and has long been a major cause
of death around the world (2). The trauma system plays an important part
in modern healthcare and has shown to lower mortality and morbidity in
injured patients through many means, such as hospital care, patient
follow up and prevention programs (3).

@@

Many hospitals have a trauma registry where they log the patients
information and timeline of what happened. The extent of the registry
varies, with HIC having more complete registries. In Sweden, 48 out of
the 49 hospitals receiving trauma patients are connected to the national
registry SweTrau, which functions as a national database designed to
enable scientific processing with epidemiological technology (20). This
logging of information is important work, as a cornerstone of trauma
quality improvement programs is multidisciplinary mortality and
morbidity (MoM) conferences. The MoM conference is a meeting where
different specialties who work within trauma care sit down and discuss
deaths and complications in order to look for preventable factors. These
conferences are performed in many hospitals globally, almost everywhere
where there are formal medical specialty departments and sometimes in
smaller hospitals as well (21). In Sweden, many hospitals have mortality
conferences on deceased patient cases, but few has MoM conferences to
better their trauma care on patients who lived (22). The reason being
that this expansion of MoM conferences takes time and resources, which
is why all hospitals don\'t implement this.

@@

The outcome is the OFI delayed treatment. An OFI is any failure in care,
e.g. clinical judgement error, any preventable or potentially
preventable death, missed diagnosis, technical error and delayed
treatment as identified and decided my the MoM conference. The outcomes
we used were \"long time until CT\" and \"long time until operation\",
which will be referred to as delayed treatment from now on.

@@

It is important to keep in mind that delay in treatment does not have a
set limit of time as a where under it is within the accepted timeframe
and if you go over it, it is considered delayed treatment. The audit
filters flag for a wait for CT over 30 minutes and a wait for operation
over 60 minutes, which then gets reviewed by previously mentioned two
specialised nurses. However, if a patient has a longer wait, it does not
necessarily mean that there was a delay. If it takes a longer time for a
trauma patient to receive a CT or operation, it might be because there
were more urgent matters who had to be taken care of, alternatively, the
MoM conference might deem it as the delay was justified for other
reasons. The point being, it is clear that the mean waiting time for CT
and operation is higher than 30 and 60 minutes respectively when looking
at the database. Using this audit filter as the definition for delay
would not mean anything, as the majority of patients would be classified
as having waited for too long. Instead, it is the MoM conferences that
decide if the delay had any adverse effects for the patient, such as an
injury worsening during the time spent waiting without a clear
motivation to why it had to wait. It is these cases that they label it
as the "too long until CT '' or "Too long until operation". Meaning, all
delays in the database are actual assessed delays, not just a time limit
that has been surpassed.

>>>>

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response (1), is and has long been a major cause
of death around the world (2). The trauma system plays an important part
in modern healthcare and has shown to lower mortality and morbidity in
injured patients through many means, such as hospital care, patient
follow up and prevention programs (3).

@@

Many hospitals have a trauma registry where they log the patients
information and timeline of what happened. The extent of the registry
varies, with HIC having more complete registries. In Sweden, 48 out of
the 49 hospitals receiving trauma patients are connected to the national
registry SweTrau, which functions as a national database designed to
enable scientific processing with epidemiological technology (20). This
logging of information is important work, as a cornerstone of trauma
quality improvement programs is multidisciplinary mortality and
morbidity (MoM) conferences. The MoM conference is a meeting where
different specialties who work within trauma care sit down and discuss
deaths and complications in order to look for preventable factors. These
conferences are performed in many hospitals globally, almost everywhere
where there are formal medical specialty departments and sometimes in
smaller hospitals as well (21). In Sweden, many hospitals have mortality
conferences on deceased patient cases, but few has MoM conferences to
better their trauma care on patients who lived (22). The reason being
that this expansion of MoM conferences takes time and resources, which
is why all hospitals don't implement this.

@@

The outcome is the OFI delayed treatment. An OFI is any failure in care,
e.g. clinical judgement error, any preventable or potentially
preventable death, missed diagnosis, technical error and delayed
treatment as identified and decided my the MoM conference. The outcomes
we used were "long time until CT" and "long time until operation", which
will be referred to as delayed treatment from now on.

@@

It is important to keep in mind that delay in treatment does not have a
set limit of time as a where under it is within the accepted timeframe
and if you go over it, it is considered delayed treatment. The audit
filters flag for a wait for CT over 30 minutes and a wait for operation
over 60 minutes, which then gets reviewed by previously mentioned two
specialised nurses. However, if a patient has a longer wait, it does not
necessarily mean that there was a delay. If it takes a longer time for a
trauma patient to receive a CT or operation, it might be because there
were more urgent matters who had to be taken care of, alternatively, the
MoM conference might deem it as the delay was justified for other
reasons. The point being, it is clear that the mean waiting time for CT
and operation is higher than 30 and 60 minutes respectively when looking
at the database. Using this audit filter as the definition for delay
would not mean anything, as the majority of patients would be classified
as having waited for too long. Instead, it is the MoM conferences that
decide if the delay had any adverse effects for the patient, such as an
injury worsening during the time spent waiting without a clear
motivation to why it had to wait. It is these cases that they label it
as the "too long until CT '' or"Too long until operation". Meaning, all
delays in the database are actual assessed delays, not just a time limit
that has been surpassed.
